---
title: "Cyclistic Bike-Share User Analysis by R"
author: "Adela Xu"
date: "2024-07-12"
output: 
  md_document: default
  pdf_document: default
  word_document: default
  html_document: default
---

# Cyclistic Bike-Share User Analysis by R

**Author:** Adela Xu  
**Date:** 2024-07-12

## **Background**

### **About the company**

**Cyclistic:** A bike-share company based in Chicago, with a fleet of 5,824 bicycles that are geotracked and locked into a network of 692 stations across Chicago. The bikes can be unlocked from one station and returned to any other station in the system anytime. There are two types of users at Cyclistic: *annual members* and *casual customers*. 

### **Business Scenario**

The company's finance analysts have concluded that annual members are much more profitable than casual riders. Thus, the marketing director (my manager) has set a clear goal: Design marketing strategies aimed at converting casual customers into annual subscribers. In order to do that, the team needs to better understand: 
<br>

* How annual subscribers and casual customers differ?
* Why casual customers would buy a subscription?
* How digital media could affect their marketing tactics?
Our team is interested in analyzing the Cyclistic historical bike trip data to identify trends.

## **Ask Phase**

### **Business Tasks**

As a data analyst on the marketing team, my job is to answer the first question: **How do annual subscribers and casual customers use Cyclistic bikes differently?** I will present a detailed analysis supported by data visualizations, share the key findings, and provide recommendations accordingly.

### **Questions to be answered**

-   What are the proportions of subscribers and casual customers?
-   How do trip durations differ for subscribers and casual customers?
-   How do trip frequencies differ between weekdays and weekends for both groups?
-   Are there noticeable differences in trip purposes between members and casual riders?
-   What are the peak hours during weekdays and weekends for both subscribers and casual customers?
-   How does bike usage vary across different seasons for both subscribers and casual customers?
-   Which stations are more popular among subscribers and casual customers?
-   How do the demographics (e.g., birth year, gender) of subscribers compare to those of casual customers?

## **Prepare Phase**
I analyzed Cyclistic's historical trip data from all four quarters of 2019 to identify trends. The data is downloaded from [here](https://divvy-tripdata.s3.amazonaws.com/index.html) and is provided by Motivate International Inc. under this [license](https://divvybikes.com/data-license-agreement). \n
Each of the four files is contains twelve columns of data for all rides that occurred between January 2019 and December 2019. The credibility of the data source is confirmed.

## **Process Phase (Cleaning and Transformation)**
Load required libraries and set global options
```{r setup-libraries}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(lubridate)
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```
Upload the four CSV files for the trip data of 2019
```{r upload-datasets}
Q1_2019 <- read.csv("Divvy_Trips_2019_Q1.csv")
Q2_2019 <- read.csv("Divvy_Trips_2019_Q2.csv")
Q3_2019 <- read.csv("Divvy_Trips_2019_Q3.csv")
Q4_2019 <- read.csv("Divvy_Trips_2019_Q4.csv")
```
Review and compare column names of each file. While the names don't have to be in the same order, we want them to match perfectly before we use a command to combine them into one file.
```{r col names}
colnames(Q1_2019)
colnames(Q2_2019)
colnames(Q3_2019)
colnames(Q4_2019)
```
Q2 dataset's column names are different from the others. Rename columns to make Q2 data consistent with the other 3 quarters'.
```{r rename col}
Q2_2019 <- rename(Q2_2019 
                  ,trip_id = X01...Rental.Details.Rental.ID
                  ,start_time = X01...Rental.Details.Local.Start.Time
                  ,end_time = X01...Rental.Details.Local.End.Time
                  ,bikeid = X01...Rental.Details.Bike.ID
                  ,tripduration = X01...Rental.Details.Duration.In.Seconds.Uncapped
                  ,from_station_id = X03...Rental.Start.Station.ID
                  ,from_station_name = X03...Rental.Start.Station.Name
                  ,to_station_id = X02...Rental.End.Station.ID
                  ,to_station_name = X02...Rental.End.Station.Name
                  ,usertype = User.Type
                  ,gender = Member.Gender
                  ,birthyear = X05...Member.Details.Member.Birthday.Year
                  )
```
Combine the four datasets into one.
```{r combine datasets}
all_trips <- bind_rows(Q1_2019, Q2_2019, Q3_2019, Q4_2019)
```
Remove any duplicate trips.
```{r remove duplicate}
all_trips <- all_trips %>% distinct()
```
Check for missing values in each column.
```{r missing values}
missing_values <- colSums(is.na(all_trips) | all_trips == "")
print(missing_values)
```
Only gender and birthyear columns have missing values.
  559206 rows with missing gender, percentage 14.6%.
  537801 rows with missing birthyear, percentage 14.1%.
Due to the high percentage of missing values in these two columns, we will retain them for further analysis that does not depend on these columns.

Check data structures of the combined dataset.
```{r all_trips str}
str(all_trips)
```
Notice：
  * start_time and end_time columns are characters. We want them to be Date-Time data for further     calculations and analysis.
  * tripduration is character. We want it to be numeric for further calculations and analysis.

Convert start_time and end_time to Date-Time.
```{r date-time_conversion}
all_trips <- all_trips %>%
  mutate(start_time = ymd_hms(start_time),
         end_time = ymd_hms(end_time))
```
Convert tripduration to Numeric.
```{r tripduration_conversion}
all_trips <- all_trips %>%
  mutate(tripduration = as.numeric(gsub(",", "", tripduration)))
```
Convert tripduration from seconds to minutes for calculation convenience.
```{r tripduration_minutes}
all_trips <- all_trips %>%
  mutate(tripduration_minutes = tripduration / 60)
```
Verify the data structure again to make sure all data are in desired data types.
```{r verify_str}
str(all_trips)
```
More data cleaning.
Delete tripduration in sec.
```{r delete_tripduration_sec}
clean_data <- all_trips %>%
  select(-tripduration)
```
Round tripduration_minutes to 2 decimal places.
```{r round_tripduration}
clean_data$tripduration_minutes <- round(clean_data$tripduration_minutes, 2)
```
Check tripduration_minutes rage
```{r tripduration_range}
range(clean_data$tripduration_minutes)
```
Notice that maximum trip duration is large. So we check how many trips are over 24h; they may be outliers and can be neglected.
```{r outliers}
sum(clean_data$tripduration_minutes >= 1440)
```
Trip over 24h percentage is 0.04%, can be treated as outliers and deleted from the dataset.
```{remove_outliers}
clean_data <- (clean_data %>%
                 filter(tripduration_minutes < 1440))
```
Check unique labels for usertype. Make sure there's only one label for each user type.
```{r unique_usertype}
unique(clean_data$usertype)
```
Check unique labels for gender. Make sure there's only one label for each gender.
```{r unique_gender}
unique(clean_data$gender)
```
See the numeric summaries of the cleaned data.
```{r summary}
summary(clean_data)
```
The data is now cleaned and transformed into desired format. Ready for further analysis!


## **Analyze and Share**
Total trip counts for both user types.
```{r total trip counts}
total_trips <- clean_data %>%
  group_by(usertype) %>%
  summarize(count = n())

total_trips$count <- format(total_trips$count, big.mark = ",")
```
Create pie chart to show total counts and proportions of both user types.
```{r pie chart}
# Format count back to number
total_trips$count_num <- as.numeric(gsub(",", "", total_trips$count))
# Calculate percentages
total_trips <- total_trips %>%
  mutate(percentage = count_num / sum(count_num) * 100)

ggplot(total_trips, aes(x = "", y = percentage, fill = usertype)) +
  geom_bar(stat = "identity") +
  coord_polar(theta = "y") +
  labs(title = "Proportion of Total Trips by User Type") +
  geom_text(aes(label = paste0(usertype, "\n", count, "\n", round(percentage, 1), "%")),
            position = position_stack(vjust = 0.5), color = "white", size = 5) + 
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = "none")
```
<br>
Create histogram to show trip duration distributions for both user types.
```{r avg trip duration by usertype}
# Calculate average trip duration for each user type
avg_tripduration <- clean_data %>%
  group_by(usertype) %>%
  summarize(avg_tripduration_minutes = round(mean(tripduration_minutes, na.rm = TRUE), 1))
```
```{r trip duration distribution}
ggplot(clean_data, aes(x = tripduration_minutes, fill = usertype)) +
  geom_histogram(binwidth = 10, boundary = 0, position = "dodge", color = "black") +
  facet_wrap(~ usertype, scales = "free_x") +
  labs(title = "Trip Duration Distribution by User Type",
       x = "Trip Duration in Minutes",
       y = "Trip Count",
       fill = "User Type") +
  xlim(0, 90) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "NONE") +
  geom_vline(data = avg_tripduration, aes(xintercept = avg_tripduration_minutes, color = usertype), linetype = "dashed", size = 1) +
  geom_text(data = avg_tripduration, aes(x = avg_tripduration_minutes, y = Inf, label = paste("Avg:", avg_tripduration_minutes, "min")),
            vjust = -0.5, hjust = 1.1, size = 3.5, color = "black", angle = 90)
```
<br>
The average trip duration for casual customers is significantly higher at 57 minutes compared to 14.3 minutes for subscribers. Casual customers typically have trip durations ranging from 10 to 30 minutes, while subscribers mostly have trip durations within the 0 to 20 minute range.

Move on to weekday/weekend analysis.
```{r add weekday col}
# add a column that shows day of the week (1 for Sunday through 7 for Saturday)
clean_data$day_of_week <- wday(clean_data$start_time, label = FALSE)
```
```{r weekday_or_weekend col}
# add a column that shows if it's weekend or weekday
clean_data$weekday_or_weekend <- ifelse(clean_data$day_of_week %in% c(1, 7), "Weekend", "Weekday")
```
```{r avg trip counts and duration for weekday/weekend}
# Calculate average trip counts and average trip duration for weekdays and weekends for subscribers and casual customers
daily_summary <- clean_data %>%
  mutate(day = as.Date(start_time)) %>%  # Extract the day from start_time
  group_by(day, weekday_or_weekend, usertype) %>%
  summarize(
    daily_trip_count = n(),
    avg_daily_tripduration_minutes = mean(tripduration_minutes, na.rm = TRUE),
    .groups = 'drop'
  )

# Calculate the average daily trip count and average daily trip duration for weekdays and weekends
average_daily_summary <- daily_summary %>%
  group_by(weekday_or_weekend, usertype) %>%
  summarize(
    avg_daily_trip_count = round(mean(daily_trip_count), 0),
    avg_daily_tripduration_minutes = round(mean(avg_daily_tripduration_minutes, na.rm = TRUE), 1),
    .groups = 'drop'
  ) %>%
  arrange(usertype, weekday_or_weekend)
```
Create a grouped bar chart to show average daily trip counts and durations by user type and day type.
```{r grouped_bar}
# Reshape the data for plotting
average_daily_summary_long <- average_daily_summary %>%
  pivot_longer(cols = c(avg_daily_trip_count, avg_daily_tripduration_minutes), 
               names_to = "metric", values_to = "value")

# Create the grouped bar chart
ggplot(average_daily_summary_long, aes(x = weekday_or_weekend, y = value, fill = usertype)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~ metric, scales = "free_y", ncol = 1) +
  geom_text(aes(label = value), 
             position = position_dodge(width = 0.9),
            vjust = 0.5, size = 4, color = "black") +
  labs(title = "Comparison of Daily Trip Counts and Durations",
       x = "",
       y = "Value",
       fill = "User Type") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```
<br>
Insights from the chart:
<br>

* Customers tend to use the bike service more on weekends compared to weekdays, indicating potential leisure needs.
* Subscribers show higher usage overall, particularly on weekdays, indicating potential commuting patterns.
* Average trip durations are generally shorter for Subscribers compared to Customers across both weekdays and weekends.

Moving on to Peak Hour analysis.
```{r}
# Extract the hours
clean_data <- clean_data %>%
  mutate(hour = hour(start_time))

# Calculate trip counts by hour, usertype, and weekday/weekend
top_trip_counts <- clean_data %>%
  group_by(hour, usertype, weekday_or_weekend) %>%
  summarize(trip_count = n()) %>%
  arrange(desc(trip_count)) %>%
  group_by(weekday_or_weekend, usertype) %>%
  top_n(2, trip_count)  # Select top 2 trips for each usertype and weekday/weekend
```
Create the histogram to show trip counts by hour for each user type and day type.
```{r peak_hour_histogram}
ggplot(clean_data, aes(x = hour, fill = usertype)) +
  geom_histogram(binwidth = 1, position = "dodge", color = "black") +
  facet_wrap(~ weekday_or_weekend + usertype, ncol = 2, scales = "free_x") +
  labs(title = "Trip Counts by Hour for Each User Type and Day Type",
       x = "Hour of Day",
       y = "Trip Count",
       fill = "User Type") +
  theme_minimal() +
  theme(strip.text = element_text(size = 12, face = "bold"),
        axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 10),
        legend.position = "none",
        strip.placement = "outside",  # Place facet labels outside the plot area
        plot.title = element_text(size = 14, face = "bold"),
        panel.spacing = unit(1, "lines")) +
  theme(axis.title.x = element_text(margin = margin(t = 20))) +  # Adjust top margin for x-axis label
  geom_text(data = top_trip_counts, aes(label = hour, y = trip_count + 2), 
            position = position_dodge(width = 1), size = 3, vjust = -0.5, color = "black")
```
<br>
Now let's do Seasonal analysis.
```{r seasonal}
# Categorize the trips into four seasons
clean_data <- clean_data %>%
  mutate(season = case_when(
    month(start_time) %in% c(12, 1, 2) ~ "Winter",
    month(start_time) %in% c(3, 4, 5) ~ "Spring",
    month(start_time) %in% c(6, 7, 8) ~ "Summer",
    month(start_time) %in% c(9, 10, 11) ~ "Fall"
  ))

# Group by season and usertype
seasonal_summary <- clean_data %>%
  mutate(season = factor(season, levels = c("Spring", "Summer", "Fall", "Winter"))) %>%
  group_by(usertype, season) %>%
  summarize(
    total_trips = n(),
    avg_trip_duration_minutes = mean(tripduration_minutes, na.rm = TRUE)
  ) %>%
  arrange(usertype, season)
```
Create the seasonal plot: a grouped bar chart to show total trips by season and user type.
```{r seasonal plot}
ggplot(seasonal_summary, aes(x = season, y = total_trips, fill = usertype)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Total Trips by Season and User Type",
    x = "Season",
    y = "Total Trips",
    fill = "User Type"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.position = "bottom",
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  )
```
<br>
Now we explore most poplular stations for both user types.
```{r top_start_stations}
# Sort the start stations by trip counts for both user types.
popular_start_stations <- clean_data %>%
  group_by(from_station_name, usertype) %>%
  summarize(trip_count = n()) %>%
  arrange(usertype, desc(trip_count))

# Filter the most popular 10 start stations for both user types.
top_start_stations <- popular_start_stations %>%
  group_by(usertype) %>%
  top_n(10, trip_count)
```
```{r top_end_stations}
# Sort the end stations by trip counts for both user types.
popular_end_stations <- clean_data %>%
  group_by(to_station_name, usertype) %>%
  summarize(trip_count = n()) %>%
  arrange(usertype, desc(trip_count))

# Filter the most popular 10 end stations for both user types.
top_end_stations <- popular_end_stations %>%
  group_by(usertype) %>%
  top_n(10, trip_count)
```
We want to combine the info in one plot. So we need to rename the columns to make them consistent.
```{r rename_stations}
top_start_stations <- top_start_stations %>%
  rename(station_name = from_station_name)

top_end_stations <- top_end_stations %>%
  rename(station_name = to_station_name)
```
Create a bubble plot to show most popular start stations and end stations for both user types.
```{r station_plot}
# Add size column
top_start_stations$size <- sqrt(top_start_stations$trip_count) * 2
top_end_stations$size <- sqrt(top_end_stations$trip_count) * 2
# Add a column for plots's label
top_start_stations$usertype_start_end <- paste(top_start_stations$usertype, "_start", sep = "")
top_end_stations$usertype_start_end <- paste(top_end_stations$usertype, "_end", sep = "")
# combine top start and end stations for combined visualization
combined_stations <- union(top_start_stations, top_end_stations)
# Create the plot
ggplot(combined_stations, aes(x = station_name, y = usertype_start_end, size = size, fill = usertype)) +
  geom_point(shape = 21, color = "black", aes(fill = usertype), stroke = 1) +
  scale_size(range = c(3, 18)) +
  labs(
    title = "Top Stations by User Type",
    x = "Station",
    y = "User Type & Start/End",
    size = "Trip Count",
    fill = "User Type"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 12),
    legend.position = "NONE",
    plot.title = element_text(size = 14, face = "bold", hjust = 0.4)
  )
```
<br>
Insight: The top start and end stations are generally consistent within the same user type. However, there is little overlap in the top stations between different user types.

Let's also do a demographic (gender & birth year) analysis.
Remember we have trip data with missing gender or birth year info. For this analysis, we need to filter out the data with missing information.
Gender analysis:
```{r gender}
# Clean the data for gender analysis.
data_with_gender <- clean_data %>%
  filter(!is.na(gender) & gender != "")

# Number and proportion of gender by User Type
gender_counts <- data_with_gender %>%
  group_by(usertype, gender) %>%
  summarise(count = n()) %>%
  ungroup()

gender_proportions <- gender_counts %>%
  group_by(usertype) %>%
  mutate(proportion = count / sum(count)) %>%
  ungroup()

# Plot proportion of genders by user type
ggplot(gender_proportions, aes(x = usertype, y = proportion, fill = gender)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_fill_manual(values = c("Male" = "blue", "Female" = "red")) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Proportion of Genders by User Type",
       x = "User Type",
       y = "Proportion",
       fill = "Gender") +
  geom_text(aes(label = paste0(gender, "\n", round(proportion, 2)*100, "%")),
            position = position_stack(vjust = 0.5), color = "white", size = 5) +
  theme(legend.position = "none",  # Remove legend
        plot.title = element_text(hjust = 0.5))  # Center plot title
```
<br>
Insight: Male customers have dominating positions in both user types. Female proportion of casual customers is slightly higher than that of subscribers.

Birth year analysis:
```{r birthyear}
# clean the data for birthyear analysis
data_with_birthyear <- clean_data %>%
  filter(!is.na(birthyear) & birthyear != "")

# Plot age distribution for Subscriber and Customer
ggplot(data_with_birthyear, aes(x = birthyear, fill = usertype)) +
  geom_histogram(binwidth = 10, boundary = 1940, position = "dodge", color = "black") +
  facet_wrap(~ usertype, ncol = 2, scales = "free_x") +
  labs(title = "Birth Year Distribution by User Type",
       x = "Birth Year",
       y = "Trip Count",
       fill = "User Type") +
  theme_minimal() +
  xlim(1940, 2010)
```
<br>
Insight: The majority of users are born between 1980 and 2000. Casual customers are predominantly born between 1990 and 2000, while subscribers are mostly born between 1980 and 1990.

## **Key Findings**

### **User Demographics**

-   Among all trips, casual customers outnumber subscribers by three to one.
-   Male users outnumber female users in both groups, but casual customers have a higher proportion of female users compared to subscribers.
-   The majority of users are born between 1980 and 2000. Casual customers are predominantly born between 1990 and 2000, while subscribers are mostly born between 1980 and 1990.

### **Trip Duration and Frequency**

-   Casual customers typically have longer trip durations (10-30 minutes) compared to subscribers (0-20 minutes).
-   Subscribers predominantly use Cyclistic on weekdays.
-   Casual users favor weekends.

### **Usage Patterns**

-   On weekdays, the peak hours for subscribers are 8 AM and 5 PM.
-   For casual customers, the peak hour on weekdays is 5 PM.
-   On weekends, the peak hours for both user types are after 12 PM.

### **Seasonal Trends**

-   Cyclistic is most popular during the summer and least popular in winter for both user types.

### **Popular Stations**

-   Popular stations differ significantly between user types.
    -   Casual customers frequent Streeter Dr & Grand Ave, Lake Shore Dr & Monroe St, and Millennium Park.
    -   Subscribers prefer Clinton St & Washington Blvd, Canal St & Adams St, and Clinton St & Madison St.

## **Act Phase**

### Recommendations

1.  **Target Younger Audiences:** Since casual users are predominantly born between 1990 and 2000, create marketing campaigns that resonate with younger demographics. Use social media platforms like Instagram, TikTok, and Snapchat where this age group is highly active.

2. **Highlight Cost Savings:** Emphasize how subscribing can save money for frequent users, especially those who currently take longer trips as casual customers. Show comparative costs between paying per trip versus subscribing for unlimited shorter trips.

3.  **Offer Flexible Subscription Plans:** Introduce flexible subscription plans that cater to varying needs, such as weekend-only plans or limited ride packages, to appeal to casual users who may not ride daily.

4.  **Weekend Specials:** Since casual users favor weekends, create special weekend promotions for subscribers. Offer perks like extended ride times or discounts on weekend rides for subscribers.

5.  **Peak Hour Benefits:** Promote benefits for subscribers during peak hours, such as guaranteed bike availability or priority access during the busiest times.

6.  **Seasonal Promotions:** Launch seasonal campaigns that offer discounts or special benefits for new subscribers, especially during the summer when usage peaks. Highlight the advantages of being a subscriber during the high-demand season.

7.  **Station-Based Incentives:** Implement incentives at popular casual user stations (Streeter Dr & Grand Ave, Lake Shore Dr & Monroe St, and Millennium Park) to encourage sign-ups. For example, set up subscription kiosks with promotional offers or host events at these locations.



